---
title: "Figure 4"
output: pdf_document
---

```{r setup, include=FALSE}
#source("/storage/mariiac/msms_figures/my_config.R")
source("../my_config.R")
library(ggbeeswarm)
```

```{r}
# Read the filtered data and removed peptides labelled contaminations
df <- read_tsv(file.path(data_path, "all_peptides_preprocessed.tsv"))
df <- df[df$is_not_contamination, ]

# Calculate miscleavages and remove >2 miscleaved peptides (TO DO!!!!)

cdr3 <- read_tsv(file.path(metadata_path, "antibodies_cdr3_variable_full.tsv"))
sample_info <- read_tsv(file.path(metadata_path, "detailed_sample_description.tsv"))

# Create combinations of replicates, for fig 4 only take the 4 merged replicates
replicate_subsets <- unlist(lapply(1:4, combn, x = paste0("run", 1:4), simplify = FALSE), recursive = FALSE)
#replicate_subsets <- replicate_subsets[15]
replicate_subsets <- replicate_subsets[1:4]
```

```{r}
res <- data.frame()
# Calculate coverage for each sample, iterate by rep combn
for (i in 1:length(replicate_subsets)) {
  for (sample in 1:70) {
    abs_all <- c("h9C12-Q97A_HC", "h9C12-WT_HC", "Brimab_HC", "PGDM1400_HC", "PGT121_HC", "Umab_HC")
    # Use only the mAb that should be in that sample as ref
    abs_ref <- abs_all[unlist(sample_info[sample, abs_all])]
    # choose the correct runs and filter for the correct sample
    subset <- replicate_subsets[[i]]
    df_tmp <- df[df$run %in% subset & df$Sample == sample, ]
    
    if (length(abs_ref)) {
      coverage <- sapply(X = abs_ref, FUN = get_coverage_percent, 
                       peptides_vector = df_tmp$Sequence, annotation = cdr3, mode = "both")
      coverage <- data.frame(t(coverage))
      coverage$subset <- paste0(subset, collapse = " ")
      coverage$sequence_name <- rownames(coverage)
      coverage$sample <- sample
      res <- rbind(res, coverage)
    }
  }
}

res$subset <- factor(res$subset,
    levels = unlist(lapply(replicate_subsets, function(x) {paste0(x, collapse = " ")})) ,ordered = TRUE)
```

```{r}
concentrations <- data.frame(read_tsv(file.path(metadata_path, "concentration_matrix.tsv")))
concentrations$sample <- 1:70
colnames(concentrations)[1:2] <- c("h9C12-Q97A", "h9C12-WT")
# Add variable for blood/non-blood
concentrations$blood <- TRUE
concentrations$blood[c(36:64,66)] <- FALSE
concentrations <- melt(concentrations, id.vars = c("sample","blood"))
concentrations$sequence_name <- paste0(concentrations$variable, "_HC")
# calculate the lvl of complexity of each sample (how many mAbs in each sample)
complexity <- concentrations %>% group_by(sample) %>% summarize(complexity = sum(value > 0))
concentrations <- left_join(concentrations, complexity, by = "sample")
concentrations$equality <- "equal"
concentrations[c(31,32,48:54,59:64),]$equality <- "unequal"

res <- merge(res, concentrations, by = c("sample", "sequence_name"), all.x = T)

res$complexity_lvl <- ifelse(res$complexity > 1, "multiple", "single")
res$sequence_name <- factor(res$sequence_name, levels = HC_names)
```

# Panel A,B: blood vs non-blood: paired plot CDRH3 coverage comparison from sample pairs of exact same input amount and composition, facet by mAb and concentration (x: blood or non-blood, y: coverage %)

```{r}
res_blood_med <- res %>% group_by(sequence_name, value, blood) %>% summarise(med_vdj = median(vdj), med_cdr3 = median(cdr3))
```

```{r fig.height=7, fig.width=5}
p_blood_vdj <- ggplot(res, aes(x = factor(blood, levels = c("TRUE", "FALSE")), y = vdj, color = blood)) +
  facet_grid(sequence_name ~ value, labeller = labeller(value = label_value)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_beeswarm(pch = 1, size = 0.8, cex = 3, corral = "wrap", corral.width = 0.9) +
  scale_color_manual(values = c("black","red3")) +
  geom_text(data = res_blood_med, aes(y = -10, label = round(med_vdj,2)), size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none",
        strip.text.x = element_blank(), strip.text.y = element_blank()
        ) +
  labs(x = "Presence of serum IgG", y = "VDJ coverage (%)")

p_blood_vdj

#ggsave("./figures/blood_vdj.pdf", p_blood_vdj, height = 7, width = 5, dpi = 600)
```

```{r fig.height=7, fig.width=5}
p_blood_cdr3 <- ggplot(res, aes(x = factor(blood, levels = c("TRUE", "FALSE")), y = cdr3, color = blood)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  facet_grid(sequence_name ~ value, labeller = labeller(value = label_value)) +
  geom_beeswarm(pch = 1, size = 0.8, cex = 3, corral = "wrap", corral.width = 0.9) +
  scale_color_manual(values = c("black","red3")) +
  geom_text(data = res_blood_med, aes(y = -10, label = round(med_cdr3,2)), size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none",
        strip.text.x = element_blank(), strip.text.y = element_blank()
        ) +
  labs(x = "Presence of serum IgG", y = "CDRH3 coverage (%)")

p_blood_cdr3

#ggsave("./figures/blood_cdrh3.pdf", p_blood_cdr3, height = 7, width = 5, dpi = 600)
```

# Panel C,D: Non-blood only, paired plot CDRH3 coverage comparison between pairs of sample of same mAb but different complexity lvl (1, 6 equal, or 6 unequal), facet by mAb and input amount (x: complexity lvl, y: coverage %).
```{r}
# Filter for non-blood samples
res_noblood <- res %>% filter(sample %in% c(36:64,66))
# Select for only h9C12WT, Umab, and PGDM1400 (only those have single mab samples)
res_noblood <- res_noblood %>% filter(sequence_name %in% c("h9C12-WT_HC", "Umab_HC", "PGDM1400_HC"))


res_noblood_med <- res_noblood %>% group_by(sequence_name, value, complexity_lvl) %>% summarise(med_vdj = median(vdj), med_cdr3 = median(cdr3))

#label_value <- c("1" = "1 ng", "10" =  "10 ng", "100" = "100 ng", "1000" = "1000 ng")
```

```{r fig.height=4.5, fig.width=5}
p_noblood_vdj <- ggplot(res_noblood, aes(x = factor(complexity_lvl, levels = c("single", "multiple")), y = vdj, color = complexity_lvl)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  facet_grid(sequence_name ~ value, labeller = labeller(value = label_value)) +
  geom_beeswarm(pch = 1, size = 1, cex = 3) +
  geom_text(data = res_noblood_med, aes(y = -10, label = round(med_vdj,2)), size = 2) +
  scale_color_manual(values = c("grey40","black")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none",
        strip.text.x = element_blank(), strip.text.y = element_blank()
        ) +
  labs(x = "Number of spike-in mAbs in sample", y = "VDJ coverage (%)")

p_noblood_vdj

#ggsave("./figures/noblood_vdj.pdf", p_noblood_vdj, height = 4.5, width = 5, dpi = 600)
```

```{r fig.height=4.5, fig.width=5}
p_noblood_cdr3 <- ggplot(res_noblood, aes(x = factor(complexity_lvl, levels = c("single", "multiple")), y = cdr3, color = complexity_lvl)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  facet_grid(sequence_name ~ value, labeller = labeller(value = label_value)) +
  geom_beeswarm(pch = 1, size = 1, cex = 3) +
  geom_text(data = res_noblood_med, aes(y = -10, label = round(med_cdr3,2)), size = 2) +
  scale_color_manual(values = c("grey40","black")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none",
        #strip.text.x = element_blank(), strip.text.y = element_blank()
        ) +
  labs(x = "Number of spike-in mAbs in sample", y = "CDRH3 coverage (%)")

p_noblood_cdr3

#ggsave("./figures/noblood_cdrh3.pdf", p_noblood_cdr3, height = 4.5, width = 5, dpi = 600)
```

# Supp figures, same thing for light chain
```{r}
res <- data.frame()

for (i in 1:length(replicate_subsets)) {
  for (sample in 1:70) {
    abs_all <- c("h9C12_LC", "Brimab_LC", "PGDM1400_LC", "PGT121_LC", "Umab_LC")
    abs_ref <- abs_all[unlist(sample_info[sample, abs_all])]
    
    subset <- replicate_subsets[[i]]
    df_tmp <- df[df$run %in% subset & df$Sample == sample, ]
    
    if (length(abs_ref)) {
      coverage <- sapply(X = abs_ref, FUN = get_coverage_percent, 
                       peptides_vector = df_tmp$Sequence, annotation = cdr3, mode = "both")
      coverage <- data.frame(t(coverage))
      coverage$subset <- paste0(subset, collapse = " ")
      coverage$sequence_name <- rownames(coverage)
      coverage$sample <- sample
      res <- rbind(res, coverage)
    }
  }
}

res$subset <- factor(res$subset,
    levels = unlist(lapply(replicate_subsets, function(x) {paste0(x, collapse = " ")})) ,ordered = TRUE)
```

```{r}
concentrations <- data.frame(read_tsv(file.path(metadata_path, "concentration_matrix.tsv")))
# For the light chain, since h9C12-WT and h9C12-Q97 are the same, take the higher value of the two cols
concentrations <- concentrations %>% mutate(h9C12 = pmax(concentrations$h9C12.Q97A, concentrations$h9C12.WT)) %>%
  select(-c(1:2)) %>% relocate(h9C12)
concentrations$sample <- 1:70
# Add variable for blood/non-blood
concentrations$blood <- TRUE
concentrations$blood[c(36:64,66)] <- FALSE
concentrations <- melt(concentrations, id.vars = c("sample","blood"))
concentrations$sequence_name <- paste0(concentrations$variable, "_LC")
# calculate the lvl of complexity of each sample (how many mAbs in each sample)
complexity <- concentrations %>% group_by(sample) %>% summarize(complexity = sum(value > 0))
concentrations <- left_join(concentrations, complexity, by = "sample")
concentrations$equality <- "equal"
concentrations[c(31,32,48:54,59:64),]$equality <- "unequal"

res <- merge(res, concentrations, by = c("sample", "sequence_name"), all.x = T)

res$complexity_lvl <- ifelse(res$complexity > 1, "multiple", "single")
res$sequence_name <- factor(res$sequence_name, levels = LC_names)
```

```{r}
res_blood_med <- res %>% group_by(sequence_name, value, blood) %>% summarise(med_vdj = median(vdj), med_cdr3 = median(cdr3))
```

```{r fig.height=5.75, fig.width=5}
ggplot(res, aes(x = factor(blood, levels = c("TRUE", "FALSE")), y = vdj, color = blood)) +
  facet_grid(sequence_name ~ value, labeller = labeller(value = label_value)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_beeswarm(pch = 1, size = 0.8, cex = 3, corral = "wrap", corral.width = 0.9) +
  scale_color_manual(values = c("black","red3")) +
  geom_text(data = res_blood_med, aes(y = -10, label = round(med_vdj,2)), size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none",
        strip.text.x = element_blank(), strip.text.y = element_blank()
        ) +
  labs(x = "Presence of serum IgG", y = "VJ coverage (%)")

#ggsave("./figures/blood_vj.pdf", height = 5.75, width = 5, dpi = 600)
```

```{r fig.height=5.75, fig.width=5}
ggplot(res, aes(x = factor(blood, levels = c("TRUE", "FALSE")), y = cdr3, color = blood)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  facet_grid(sequence_name ~ value, labeller = labeller(value = label_value)) +
  geom_beeswarm(pch = 1, size = 0.8, cex = 3, corral = "wrap", corral.width = 0.9) +
  scale_color_manual(values = c("black","red3")) +
  geom_text(data = res_blood_med, aes(y = -10, label = round(med_cdr3,2)), size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none",
        strip.text.x = element_blank(), strip.text.y = element_blank()
        ) +
  labs(x = "Presence of serum IgG", y = "CDRL3 coverage (%)")

#ggsave("./figures/blood_cdrl3.pdf", height = 5.75, width = 5, dpi = 600)
```

```{r}
# Filter for non-blood samples
res_noblood <- res %>% filter(sample %in% c(36:64,66))
# Select for only h9C12WT, Umab, and PGDM1400 (only those have single mab samples)
res_noblood <- res_noblood %>% filter(sequence_name %in% c("h9C12_LC", "Umab_LC", "PGDM1400_LC"))

res_noblood_med <- res_noblood %>% group_by(sequence_name, value, complexity_lvl) %>% summarise(med_vdj = median(vdj), med_cdr3 = median(cdr3))
```

```{r fig.height=4.5, fig.width=5}
ggplot(res_noblood, aes(x = factor(complexity_lvl, levels = c("single", "multiple")), y = vdj, color = complexity_lvl)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  facet_grid(sequence_name ~ value, labeller = labeller(value = label_value)) +
  geom_beeswarm(pch = 1, size = 1, cex = 3) +
  geom_text(data = res_noblood_med, aes(y = -10, label = round(med_vdj,2)), size = 2) +
  scale_color_manual(values = c("grey40","black")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none",
        strip.text.x = element_blank(), strip.text.y = element_blank()
        ) +
  labs(x = "Number of spike-in mAbs in sample", y = "VJ coverage (%)")

#ggsave("./figures/noblood_vj.pdf", height = 4.5, width = 5, dpi = 600)
```

```{r fig.height=4.5, fig.width=5}
ggplot(res_noblood, aes(x = factor(complexity_lvl, levels = c("single", "multiple")), y = cdr3, color = complexity_lvl)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  facet_grid(sequence_name ~ value, labeller = labeller(value = label_value)) +
  geom_beeswarm(pch = 1, size = 1, cex = 3) +
  geom_text(data = res_noblood_med, aes(y = -10, label = round(med_cdr3,2)), size = 2) +
  scale_color_manual(values = c("grey40","black")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none",
        strip.text.x = element_blank(), strip.text.y = element_blank()
        ) +
  labs(x = "Number of spike-in mAbs in sample", y = "CDRL3 coverage (%)")

#ggsave("./figures/noblood_cdrl3.pdf", height = 4.5, width = 5, dpi = 600)
```